Fully convective star that produced a near-polar flare shows a typical corona

### 2023-08-09 src/tex/ms.tex
[showyourwork] overleaf sync

% Title
\title{Does a fully convective star need a special corona to produce a polar flare?}

% Author list
\author{Ekaterina Ilin, Katja Poppenh\"ager, Beate Stelzer}

%\affiliation
\author[0000-0002-6299-7542]{Ekaterina Ilin}
\affiliation{Leibniz Insitute for Astrophysics Potsdam (AIP), An der Sternwarte 16, 14482 Potsdam, Germany}

@@

% Abstract with filler text
\begin{abstract}
    TIC 277 is the star with the highest latitude flare located to date. In 2020, TESS observed this rapidly rotating M7 dwarf produce a flare at 81 deg latitude. This is in stark contrast to solar flares that occur much closer to the equator, typically below 30 deg. Does a star require a special corona to produce polar flares?

    We studied five Sectors of TESS observations, and obtained 36 ks ob XMM-Newton observations, to find that TIC 277's corona does not differ significantly from other low mass stars on the saturated activity branch with respect to X-ray luminosity, coronal temperatures, and flaring activity. Additionally, XMM-Newton captured a simultaneous X-ray and optical flare on TIC 277 with a bolometric energy of \eom and a soft X-ray energy of \eepic in the 0.2-2 keV band. Combining the 18 flares found in TESS with the flare from XMM-Newton, we inferred that TIC 277 produces flares at or above this energy approximately once every 10 h. We suggest that TIC 277's X-ray and flaring activity is representative of other rapidly rotating late M dwarfs, but that further measurements of localized flares are needed to understand the conditions under which high latitude flares in these stars can occur.

@@

% Main body with filler text
\section{Introduction}
\label{sec:intro}
The smallest stars are crawling with rocky planets. The disk mass around the smallest stars decreases such that giant planets are no longer efficiently formed, leaving room for smaller objects in the orbit~(XXXX). Terrestrial planets in the habitable zones of these stars are a factor of XXXX more likely than around higher mass stars~\citep{hardegree-ullman2019kepler}. However, to be habitable in an Earth-like sense, merely the possibility of surface water in its liquid form does not suffice. Space weather of the host, that is, the high energy radiation and particles that force the planet's atmosphere, is another important factor. If the energetic particle flux is too high, the atmosphere may blow off. If not energetic enough, life may not emerge in the first place.

At the lowest masses, a star's evolution unfolds much slower than for a Sun. While the Sun's magnetic activity, and with it the flares, winds, coronal mass ejections and energetic particle eruption that make up space weather, decline rapidly over a few hundred megayears, a fully convective M dwarf stays highly active for gigayears. Add that the zone where liquid water can reside is much closer to these faint stars, and the result is a planet that may be exposed to violent space weather conditions for a good fraction of the age of the universe.

Space weather originates from the stellar corona. The transition from star to brown dwarf is characterized by increasing rotation speed and decreasing coronal emission. The simultaneous surge of radio emission indicates a transition from stellar corona to planet-like magnetosphere at the bottom of the main sequence. Yet, brown dwarfs are regularly detected with energetic flares regardless of their low X-ray luminosity. How these flares are produced in the apparent absence of a solar-type corona is unclear. An intermittent corona is conceivable, similarly XYZ. Perhaps, the rapid rotation of these low-mass objects plays a role, as it governs the dynamo process within, and consequently influences the complex process of flux emergence~\citep{weber2023understanding}.

\citet{ilin2021giant} searched fully convective dwarfs observed in TESS, and found four flares on four rapidly rotating ($P<10\,$h) M5-M7 dwarfs. These flares' durations spanned multiple rotation periods of each star, so that the flares were observed to rotate in and out view. The shape of this modulation together with known inclination (combining $P$ and $v \sin i$ from high resolution spectroscopy) allowed them to determine the latitudes of these flares. Surprisingly, these flares were found closer to the pole than to the equator, in contrast to the Sun, where flares are usually found below 30 deg latitude. Their results indicate a preference for flares at high latitudes in these star, as a chance finding was unlikely ($0.1\%$). If flares and associated particle eruptions have a latitude preference, the space weather of a planet in an aligned orbit would be less severe than previously thought. %Are high flaring latitudes  an effect of the dynamo at high rotation speeds? Or is there something special about these stars' coronae that allowed them to produce flares at these latitudes?

@@

TIC 277539431, TIC 277 for short, is one out of four stars with a high-latitude flare in its TESS observations~\citep{ilin2021giant}. In fact, it showed the highest latitude flare known to date at $81\pm1\,$deg. TIC 277 is an M7 dwarf with a rotation period of 4.56 hours. Its age is unknown. \citet{schneider2018hazmat} classify it as a field star, but its FUV and NUV fluxes are on the high end for field stars and would be compatible with the NUV/FUV emission from younger open cluster members, as well. 
Is its high flaring latitude an effect of the dynamo at high rotation speeds? Or is there something special about this star's corona that led it to produce are flare at these latitudes?

In this work, we follow TIC 277 with XMM-Newton for 36 ks and use optical monitoring from the Transiting Exoplanet Survey Satellite~(TESS,~\citealt{ricker2015transiting}) to measure its coronal properties, and investigate if producing a high-latitude flare requires an unusual corona, or if it is rather typical of stars of this spectral type and rotation.

@@

    \end{tabular}
    \newline\footnotesize
    [1] \citet{ilin2021giant}, 
    [2] \citet{bailer-jones2018estimating}, 
    % [3] 2MASS, \citet{skrutskie2006two}, 
    % [4] \citet{collaboration2016gaia,gaiacollaboration2023gaia}, 
    [3] \citet{pecaut2013intrinsic}
    \label{tab:modelparams}
\end{table}

\section{Observations}

@@

For the OM flare, we derived the bolometric flare energy following the procedure in~\citet{shibayama2013superflares}. We used the throughput curve for the white light filter scaled to unity at the peak of transmission as an optimistic response curve $R_{\lambda, \rm OM}$, given that the degradation of this filter's response is poorly constrained. From that, we extracted a ratio $f_{\rm OM}$ of stellar to flare flux of $1.4\cdot 10^{-4}$ for a $10,000\,$K blackbody flare\footnote{A flare temperature of $10,000\,$K is a typical approximation for energetic M dwarf flares~\citep{kowalski2013timeresolved, howard2020evryflarea}.}:

@@

We note that due to the optimistic response curve assumption the resulting $E_{\rm flare}$ is likely higher in reality.

@@

\section{Results}

With the XMM-Newton and TESS observations, we could constrain the coronal and flaring properties of TIC 277. We derive its coronal temperature and luminosity from the EPIC spectrum~(Section~\ref{sec:res:XrayTL}). We also measure the flare energies in X-ray from the EPIC instruments, and optical from the OM, and compare them to the flare frequency distributions obtained from the 18 detected flares in TESS~(Section~\ref{sec:res:flares}).

@@

\subsection{Coronal temperature}
\label{sec:discussion:xraytemp}
Low mass stars' coronae are not uniform. Magnetic structures create a range of temperatures, starting at about 1 MK by definition and rising up to 30 MK in flares, in the heterogeneous Solar corona~\citep{vaiana1978recent}. Stellar point source X-ray observations are a superposition of various temperatures and densities. The resolution of different temperature component depends on the apparent brightness and integration time of the spectrum. Nearby stars like Proxima Cen can be modeled with a range of temperatures without overfitting~\citep[e.g.,][]{gudel2004flares}, while one can often only capture one dominant temperature in fainter objects~(XXXX).

@@

\subsection{X-ray luminosity}
\label{sec:discussion:xraylum}
Similar to its temperature make-up, while TIC~277's coronal luminosity relative to bolometric luminosity of $L_X/L_{\rm bol}=$\LXLbol is somewhat lower than the $L_X/L_{\rm bol}\sim 10^{-3}$ average in the saturated regime~\citep{wright2011stellaractivityrotation,wright2018stellar}, it is not an outlier to other M dwarfs in this activity regime.

In the Hyades and Praesepe, \citet{nunez2022factory} cover both fully convective and rapidly rotating M dwarfs, and find a downward trend in $L_X/L_{\rm bol}$ with decreasing rotation period for M3.5-M7 dwarfs, tentatively pointing toward a supersaturation of the corona at fast rotation speeds, as has been conjectured before~\citep{jeffries2011investigating,ramsay2020tess}.   \citet{jeffries2011investigating} suggest that the cause for supersaturation is centrifugal stripping of the corona, but \citet{reiners2022magnetism} show that if the averae surface magnetic field strength increases, $L_X/L_{\rm bol}$ increases likewise with no sign of a break, up to $10^{-3}$. They conclude that the trend is rather due to the underlying dynamo. If TIC~277 is indeed underluminous, these results suggest that its dynamo produces a weaker magnetic field than that of other saturated M dwarfs, which results in a slightly lower X-ray luminosity. However, since the flaring activity of TIC~277 is in line with saturated activity, this conjecture currently lacks evidence from other activity indicators~\citep[see][, for supersaturation in flaring activity in fully convective M dwarfs]{ramsay2020tess}. Moreover, late M dwarfs are underrepresented in these studies.

\citet{stelzer2022first} investigated late M and L dwarfs with eROSITA to show that $L_X/L_{\rm bol}$ increases from the full convection boundary (M3) towards later spectral types, but begins to decrease again from M7 downward. In that study, TIC~277 is at the lower boundary of the $L_X/L_{\rm bol}$ distribution for M7 dwarfs. At the same time, strong magnetic fields are found across the board in brown dwarfs~(XXXX). While it is not clear whether M7 falls in the stellar regime of~\citet{reiners2022magnetism} or into that latter brown dwarf realm, or has features of both, it appears at the lower bound regardless. The small number of supersaturated stars in \citet{nunez2022factory}  can only tentatively suggest that this is a common phenomenon among mid-to-late M dwarfs when they are quickly rotating.

@@

\section{Summary and Conclusions}
We investigated whether TIC 277, a rapidly rotating M7 dwarf with a flare localized at 81 deg latitude, shows exceptional coronal or flaring properties to produce said flare so close to the rotational pole. We obtained 36 ks of EPIC and OM observations with XMM-Newton, and studied the five Sectors of red-optical 27-day light curves provided by TESS. We found an X-ray luminosity of XXXX, and a mean coronal temperature of XXXX MK, typical of rapidly rotating late M dwarfs. The flaring activity is also typical of late M dwarfs in the saturated regime, with a flare rate $f(> log 10 Eflare = 31.5 erg)\approx -1.1$.

XMM-Newton also captured a simultaneous X-ray and optical flare with its EPIC and OM instruments, respectively. The flare energy and rate of the OM flare are consistent with the flare frequency distribution of the 18 detected TESS flares, indicating that X-ray flares on TIC 277 can be observed with XMM-Newton about once every $10\,$h. The bolometric and soft-X-ray flare energies are within an order of magnitude within each other, which places them among other stars with simultaneous X-ray and optical flare energies~\citep{guarcello2019simultaneous, kuznetsov2021stellar}, as well as the Sun~(XXXX).

While TIC 277's coronal and flaring properties do not appear uncommon, this does not mean that the corona has no influence on the high latitude occurrence of flares. TIC 277 is a fully convective rapid rotator with a rotation period of only $4.56\,$h. The detected high-latitude flare in TESS~\citep{ilin2021giant} may hence be a product of high-latitude flux emergence driven by its fast rotation~\citep{weber2016modeling,weber2017suppression}.
However, the type of magnetic dynamo driven by such fast rotation may produce a peculiarly structured corona~(XXXX), which then appears common within the mass-rotation regime of TIC 277. Less than ten flare latitudes on stars other than the Sun are known to date~\citep{wolter2008doppler, ilin2021giant, johnson2021simultaneous}. Localization of more flares on the surfaces of a broader range of stars will allow us to discriminate the role of dynamo and flux emergence theories from that of coronal properties in the production of active latitudes.

>>>>

% Title
\title{Fully convective star that produced a near-polar flare shows a typical corona}

% Author list with  \affiliation

\author[0000-0002-6299-7542]{Ekaterina Ilin}
\affiliation{Leibniz Insitute for Astrophysics Potsdam (AIP), An der Sternwarte 16, 14482 Potsdam, Germany}

@@

% Abstract with filler text
\begin{abstract}
    In 2020, TESS observed a rapidly rotating M7 dwarf produce a flare at 81 deg latitude, the highest latitude flare located to date. This is in stark contrast to solar flares that occur much closer to the equator, typically below 30 deg. The circumstances that allow flares at high latitudes to occur are poorly understood. We studied five Sectors of TESS observations, and obtained 36 ks of XMM-Newton observations, to find that TIC 277's corona does not differ significantly from other low mass stars on the saturated activity branch with respect to X-ray luminosity, coronal temperatures, and flaring activity. Additionally, XMM-Newton captured a simultaneous X-ray and optical flare on TIC 277 with a bolometric energy of \eom and a soft X-ray energy of \eepic in the 0.2-2 keV band. Combining the 18 flares found in TESS with the flare from XMM-Newton, we inferred that TIC 277 produces flares at or above this energy approximately once every 10 h. We conclude that TIC 277's X-ray and flaring activity is representative of other rapidly rotating late M dwarfs, but that further measurements of localized flares are needed to understand the conditions under which high latitude flares in these stars can occur.

@@

% Main body with filler text
\section{Introduction}
\label{sec:intro}
Most M dwarfs hosts terrestrial planets with a sizable fraction orbiting at instellations where liquid water can exist~\citep{dressing2015occurrence, hardegree-ullman2019kepler, ment2023occurrence}. However, to be habitable in an Earth-like manner, merely the possibility of surface water in its liquid form does not suffice. Space weather of the host, that is, the high energy radiation and particles that force the planet's atmosphere affects hospitability for life, for M dwarfs in particular~\citep{airapetian2020impact}. If the energetic (particle) flux is too high, the atmosphere may blow off~(XXXX). If not energetic enough, life may not emerge in the first place~\citep{rimmer2018origin}.

In M dwarfs, stellar activity evolution on the main sequence unfolds much slower than for a FGK star. While for the latter, magnetic activity, and with it the flares, winds, coronal mass ejections and energetic particle eruptions that make up their space weather, decline rapidly over a few hundred megayears, a fully convective M dwarf stays highly active for gigayears~\citep{XXXX}. Moreover, the zone where liquid water can reside is much closer to these faint stars. The result is a planet that may be exposed to violent space weather conditions for a good fraction of the age of the universe.

Stellar space weather originates from the stellar corona. The transition from star to brown dwarf is characterized by increasing rotation speed and decreasing coronal emission. The simultaneous surge of radio emission indicates a transition from stellar corona to planet-like magnetosphere at the bottom of the main sequence~(XXXX). Yet, brown dwarfs are regularly detected with energetic flares regardless of their low X-ray luminosity~(XXXX). How these flares are produced in the apparent absence of a solar-type corona is unclear~\citep{weber2023understanding}.%. An intermittent corona is conceivable, similarly XXXX. Perhaps, the rapid rotation of these low-mass objects plays a role, as it governs the dynamo process within, and consequently influences the complex process of flux emergence

\citet{ilin2021giant} searched fully convective dwarfs that straddle the transition from star to brown dwarf with TESS. They found four flares on four rapidly rotating ($P<10\,$h) M5-M7 dwarfs that lasted for multiple rotation periods of each star. The flares were observed to rotate in and out view. The shape of this modulation together with known inclination (combining $P$ and $v \sin i$ from high resolution spectroscopy) allowed them to determine the latitudes of these flares. Surprisingly, these flares were found significantly closer to the pole than to the equator, in contrast to the Sun, where flares are usually found below 30 deg latitude. Their results indicate a preference for flares at high latitudes in these stars, as a chance finding was unlikely ($\sim 0.1\%$). If flares and associated particle eruptions indeed have a latitude preference, the space weather of a planet in an aligned orbit would be less severe than previously thought. %Are high flaring latitudes  an effect of the dynamo at high rotation speeds? Or is there something special about these stars' coronae that allowed them to produce flares at these latitudes?

@@

TIC 277539431, TIC 277 for short, is one out of four stars with a high-latitude flare in its TESS observations~\citep[][Table \ref{tab:starparams}]{ilin2021giant}. In fact, it showed the highest latitude flare known to date at $\sim81\,$deg. TIC 277 is an M7 dwarf with an extremely short rotation period of 4.56 hours, detected in the rotational modulation of its light curve. While the fast rotation and late spectral type could be indicative of a brown dwarf~(), the detection of flares suggests the presence of a stellar corona~().%The star's age is unknown. \citet{schneider2018hazmat} classify it as a field star, but its FUV and NUV fluxes are on the high end for field stars and would, if not as high, be compatible with the NUV/FUV emission from younger open cluster members. 
%Is its high flaring latitude an effect of the dynamo at high rotation speeds? Or is there something special about this star's corona that led it to produce are flare at these latitudes? 
In this work, we follow TIC 277 with XMM-Newton for 36 ks and use optical monitoring from the Transiting Exoplanet Survey Satellite~(TESS,~\citealt{ricker2015transiting}, Section~\ref{sec:obs}) to measure its coronal and flaring properties, respectively~(Section~\ref{sec:methods}). We contrast our results~(Section~\ref{sec:results}) with the literature to investigate if TIC~277 behaves like a low mass star, brown dwarf, or something in between~(Section~\ref{sec:discussion}), and conclude in Section~\ref{sec:summary}.

%producing a high-latitude flare requires an unusual corona, or if it is rather typical of stars of this spectral type and rotation.

@@

    \end{tabular}
    \newline\footnotesize
    [1] \citet{ilin2021giant}, 
    [2] \citet{bailer-jones2018estimating}, 
    % [3] 2MASS, \citet{skrutskie2006two}, 
    % [4] \citet{collaboration2016gaia,gaiacollaboration2023gaia}, 
    [3] \citet{pecaut2013intrinsic}
    \label{tab:starparams}
\end{table}

\section{Observations}
\label{sec:obs}

@@

For the OM flare, we derived the bolometric flare energy following the procedure in~\citet{shibayama2013superflares}. We used the throughput curve for the white light filter scaled to unity at the peak of transmission as an optimistic response curve $R_{\lambda, \rm OM}$, given that the degradation of this filter's response is poorly constrained\footnote{XMM-Newton User's Handbook 3.5.3.1, \url{https://xmm-tools.cosmos.esa.int/external/xmm_user_support/documentation/uhb/omfilters.html}, accessed Aug 8, 2023}. From that, we extracted a ratio $f_{\rm OM}$ of stellar to flare flux of $1.4\cdot 10^{-4}$ for a $10,000\,$K blackbody flare\footnote{A flare temperature of $10,000\,$K is a typical approximation for energetic M dwarf flares~\citep{kowalski2013timeresolved, howard2020evryflarea}.}:

@@

We note that due to the optimistic response curve assumption, the resulting $E_{\rm flare}$ is likely higher in reality.

@@

\section{Results}
\label{sec:results}
With the XMM-Newton and TESS observations, we could constrain the coronal and flaring properties of TIC 277. We derive its coronal temperature and luminosity from the EPIC spectrum~(Section~\ref{sec:res:XrayTL}). We also measure the flare energies in X-ray from the EPIC instruments, and optical from the OM, and compare them to the flare frequency distributions obtained from the 18 detected flares in TESS~(Section~\ref{sec:res:flares}).

@@

\subsection{Coronal temperature}
\label{sec:discussion:xraytemp}
Low mass stars' coronae are not uniform. Magnetic structures create a range of temperatures, starting at about 1 MK by definition and rising up to 30 MK in flares, in the heterogeneous Solar corona~\citep{vaiana1978recent}. Stellar point source X-ray observations are a superposition of various temperatures and densities. The resolution of different temperature component depends on the apparent brightness and integration time of the spectrum. Nearby stars like Proxima Cen can be modeled with a range of temperatures without overfitting~\citep[e.g.,][]{gudel2004flares, drake2020pointing}, while one can often only capture one dominant temperature in fainter objects~(XXXX).

@@

\subsection{X-ray luminosity}
\label{sec:discussion:xraylum}
TIC~277's coronal luminosity relative to bolometric luminosity of $L_X/L_{\rm bol}=$\LXLbol in quiescence is about an order of magnitude lower than the $L_X/L_{\rm bol}\sim 10^{-3}$ average in the saturated regime of partly and fully convective M dwarfs~\citep{wright2011stellaractivityrotation,wright2018stellar}. In the Hyades and Praesepe, \citet{nunez2022factory} cover both fully convective and rapidly rotating M dwarfs, and find a downward trend in $L_X/L_{\rm bol}$ with decreasing rotation period for M3.5-M7 dwarfs. This tentatively points toward a supersaturation of the corona at fast rotation speeds, as has been suggested before for M dwarfs~\citep{jeffries2011investigating}, but also earlier type stars~\citep{prosser1996rosat, argiroffi2016supersaturation} and even red giants~\citep{dixon2020rotationally}. None of the M dwarf studies probe below spectral type M6. At later spectral type, a decline in coronal activity is expected as the magnetosphere transitions from a stellar to a planetary one~\citep{pineda2017panchromatic}. \citet{stelzer2022first} investigated late M and L dwarfs with eROSITA to show that $L_X/L_{\rm bol}$ increases from the full convection boundary (M3) towards later spectral types~\citep{magaudda2022firsta}, but begins to decrease again from M7 downward. In a compilation together with previous work from \citet{stelzer2012, cook2014trends, deluca2020extras, williams2014trends}, TIC~277 appears at the lower boundary of the $L_X/L_{\rm bol}$ distribution for M7 dwarfs. However, the rotation periods are unknown for most of these late M dwarfs. Sometime projected rotation periods are known~\citep{cook2014trends}, but true periods are only detected in a handful of these stars: TRAPPIST-1 is similar to TIC~277 with $L_X/L_{\rm bol}\sim 10^{-3}$ at its slower 1.4 d rotation period.

At the same time, strong magnetic fields are found across the board in brown dwarfs~(XXXX).

If TIC~277 is indeed systematically underluminous, these results suggest that its dynamo produces a weaker magnetic field than that of other saturated M dwarfs, which results in a slightly lower X-ray luminosity. However, since the flaring activity of TIC~277 is in line with saturated activity, this conjecture currently lacks evidence from other activity indicators~\citep[see][, for supersaturation in flaring activity in fully convective M dwarfs]{ramsay2020tess}. As often, late M dwarfs are underrepresented in these studies.

% While it is not clear whether M7 falls in the stellar regime of~\citet{reiners2022magnetism} or into that latter brown dwarf realm, or has features of both, it appears at the lower bound regardless. The small number of supersaturated stars in \citet{nunez2022factory}  can only tentatively suggest that this is a common phenomenon among mid-to-late M dwarfs when they are quickly rotating.

@@

\section{Summary and Conclusions}
\label{sec:summary}
We investigated whether TIC 277, a rapidly rotating M7 dwarf with a flare localized at 81 deg latitude, shows exceptional coronal or flaring properties to produce this flare so close to the rotational pole. We obtained 36 ks of EPIC and OM observations with XMM-Newton, and studied the five Sectors of red-optical 27-day light curves provided by TESS. We found an X-ray luminosity of \LX, and a mean quiescent coronal temperature of \Tqmean, typical of rapidly rotating fully convective M dwarfs. The flaring activity is also typical of late M dwarfs in the saturated regime, with a flare rate $f(> log 10 Eflare = 31.5 erg)\approx$\ffdr.

XMM-Newton also captured a simultaneous X-ray and optical flare with its EPIC and OM instruments, respectively. The flare energy and rate of the OM flare are consistent with the flare frequency distribution of the 18 detected TESS flares, indicating that X-ray flares on TIC 277 can be observed with XMM-Newton about once every $10\,$h. The bolometric and soft-X-ray flare energies are within an order of magnitude within each other, which places them among other stars with simultaneous X-ray and optical flare detections, although, typically, the white light emission dominates over soft X-ray~\citep{guarcello2019simultaneous, kuznetsov2021stellar}.

While TIC 277's overall coronal and flaring properties do not appear uncommon, this does not mean that all activity saturated fully convective dwarfs produce high-latitude flares. TIC 277 is an extreme rapid rotator with a rotation period of only $4.56\,$h. The detected high-latitude flare in TESS~\citep{ilin2021giant} may hence be a product of high-latitude flux emergence driven by its fast rotation~\citep{weber2016modeling,weber2017suppression}. This process may in turn produce a spatially structured corona that is not resolved in our observations, so that it appears typical of slower rotators on the saturated activity branch. Less than ten flare latitudes on stars other than the Sun are known to date~\citep{wolter2008doppler, ilin2021giant, johnson2021simultaneous}. Localization of more flares on the surfaces of a broader range of stars will allow us to discriminate the role of dynamo and flux emergence theories from that of coronal properties in the production of active latitudes.
